Red-Giant Hot Jupiters: Brilliant Radio Beacons?

### 2015-09-03 ms.tex
filename of cartoon figure is modified
response to Dave about LWA

@@

Another ingredient needed for modeling the radio emission from the interaction between the planetary magnetic field and the stellar wind is the density and velocity of the stellar wind. 
The number density of \memoTM{electrons in the (?)} stellar wind, $n$, can be expressed as
%%%%%%%%%% 
\begin{equation}
n = \frac{\dot M_\star}{4\pi a^2 m_p v}
\label{eq:n}
\end{equation}
%%%%%%%%%% 
where $\dot M_\star$ is the stellar mass loss rate, $a$ is the orbital distance from the star and $m_p$ is proton mass, and $v$ is the velocity of the stellar wind.

@@

We estimate the radio emission of exoplanets, $P_{\rm radio}$, simply by scaling the Jovian auroral radio emission, $P_{\rm radio,\,J}$, with the input energy from stellar wind, in the same manner as \citet{griesmeier2005,griesmeier2007a,griesmeier2007b}.
The scaling is based on the empirical/apparent good correlation between the radio emission intensity of Solar System planets and the input kinetic energy, $P_{\rm inp,\,k}$, or the magnetic energy of the solar wind, $P_{\rm inp,\,m}$, \citep[``radio Bode's law''; ][]{desch+kaiser1984}, i.e.,
%Namely, although the detail of the mechanism to generate planetary radio emission is not fully understood so far \memoYF{cite something?}, 
%observations indicate that the total radio emission, $P_{\rm radio}$ is proportional to the input kinetic energy, $P_{\rm k,\,inp}$, or the input magnetic energy, $P_{\rm m,\,inp}$ \citep[``radio Bode's law''; ][]{desch+kaiser1984}:
%%%
\begin{eqnarray}
P_{\rm radio} &\propto & P_{\rm inp} \\
P_{\rm inp,\,k} &=& n v^3 r_s ^2 \label{eq:Pinp_kin}, \\
P_{\rm inp,\,m} &=& v B_{\bot }^2 r_s ^2 \label{eq:Pinp_mag} \, ,
\end{eqnarray}
\memoTM{Some kind of plot of the typical ranges of these quantities might be useful/illustrative.}
%%%
where $ B_{\bot }$ is the interstellar magnetic field perpendicular to the stellar wind flow, and $r_s$ is the radius of the magnetic stand-off point. 
%In the case of the Solar wind, the kinetic energy is dominant input energy over the magnetic one (the former is $\sim 400$ times larger than the latter).  
It is not clear from the observations of the Solar System planets which is the fundamental one, the correlation to the input kinetic energy, that to the magnetic energy, or both. 
In this paper, we assume that the radio emission is scaled with the input kinetic energy, i.e., $P_{\rm radio} \propto P_{\rm k,\,inp}$ and consider the possible effects of massive stellar wind.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{deluxetable*}{cccccc}
\tabletypesize{\footnotesize}
\tablecolumns{6}
\tablecaption{Detectability of Hot Jupiters with Current 
and Future Radio Telescopes.\label{tab:sens}}
\tablehead{
Instrument & Band  & $\sigma_{\rm RMS}$\tablenotemark{a} & $\tau$\tablenotemark{b}  & Resolution & $\sigma_c$\tablenotemark{c}\\
           & (MHz) & ($\mu$Jy~hr$^{1/2}$)         &  (hr)                    & (\arcsec)  & ($\mu$Jy bm$^{-1}$)} 
\startdata
LOFAR-LBA  &  30-80  & 700   & 120000             & 8.3  & 42 \\ 
LOFAR-HBA  & 110-200 &  35   &    310             & 3.3  & 1.1 \\  
HERA       &  30-80  & 870   & 160000             & 1030 & 4.2$\times10^5$\\ 
GMRT       & 130-190 & 430   &  46000             & 16.5 & 56.9 \\ 
%LWA       &  30-80  & 58000 &  702127141         & 10.3 & 42.2 \\
%MWA       &  30-80  & 29000 &  176223211         & 1030 & 4.2$\times10^5$ \\
SKA-Low (part) & 50-200 & 7.8 &  12.6             & 7.6  & 9.7 \\
SKA-Low (full) & 50-350 & 2.2 &  1.0              & 4.8  & 1.5 \\
%VLA 4-Band& 30-80   & 3300  & 2312058            & 28.6 & 330 \\
VLA-LOBO   &  50-350 &  92   &  1800              & 8.6  & 10.4
\enddata
%\tablecomments{Time estimates to reach 2~$\mu$Jy RMS.}
\tablenotetext{a}{RMS noise in a 1-hour observation, computed for imaging of the full band reported 
in this table (Equation~\ref{eq:rms_sens}).}
\tablenotetext{b}{The integration time in hours required to reach an RMS of 2~$\mu$Jy (i.e.\ 5-$\sigma$ detection
of a typical source at 100~pc with 10~$\mu$Jy bm$^{-1}$ average flux).}
\tablenotetext{c}{Source confusion RMS contribution.}
\end{deluxetable*}
%
% HERA
% http://reionization.org/wp-content/uploads/2015/05/HERA0_imaging.pdf
%
% LOFAR
% https://www.astron.nl/~heald/test/sens.php; 
% 24 core stations, 16 remote, 8 international, 488 subbands
%
% GMRT
% http://gmrt.ncra.tifr.res.in/~astrosupp/obs_setup/sensitivity.html
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The primary limitation to detectability is a telescope's sensitivity, characterized by the root mean square (RMS) noise fluctuations in the sky and receiver.
The RMS noise ${\sigma}_{\rm RMS}$ in an interferometric observation with integration time $\tau$, bandwidth $\Delta \nu$, effective area $A_{\rm eff}$ ($\approx0.7 \pi (D/2)^2$ for interferometers comprising antennas with dish size $D$), and $N$ antennas is \citep[see e.g.][]{condon+ransom2016}
%%%
\begin{equation}
\sigma_{\rm RMS} = \frac{2 k T_{\rm sys}}{A_{\rm eff} \, \sqrt{\Delta \nu \, \tau N (N-1)}} 
\left(\frac{\rm 10^{26} Jy}{\rm W m^{-2} Hz^{-1}}\right).
\label{eq:rms_sens}
\end{equation}
%%%
Here, $T_{\rm sys}$ is the blackbody temperature equivalent of the total system noise, which is the sum of the instrument or receiver noise temperature $T_{\rm rx}$ and the noise contribution from the sky $T_{\rm sky}$.
We estimate $T_{\rm sky}$ using the low frequency sky noise temperature fit from \cite{Rogers+Bowman2008}:
%%%
\begin{equation}
T_{\rm sky} = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + T_{\rm CMB} 
\end{equation}
%%%
where $T_{150} = 283.2~\rm K$ is the $T_{\rm sky}$ at 150~MHz, $\beta=2.47$, and $T_{\rm CMB} = 2.73~\rm K$ is the contribution from the cosmic microwave background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors found it to be consistent with published measurements from 10-408~MHz. 
%The $T_{\rm rx}$ is the receiver noise temperature, which depends on the instruments and is described below.

@@

% LOFAR
The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their high and low bands, differing configurations, and antenna designs, which are called respectively the Low Band Antennas (LBA) and High Band Antennas (HBA).
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 100-240~MHz range \citep{vanHaarlem2013}.
We consider it as two separate instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and LOFAR-HBA's is at 150~MHz.
Given the strong frequency dependence of the effective aperture of a dipole antenna, and that the system noise temperature $T_{\rm sys}$ for LOFAR is not fully determined by the sky noise $T_{\rm sky}$ (i.e.\ $T_{\rm rx} \sim T_{\rm sky}$), we use the LOFAR Image Noise calculator\footnote{Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which is based on {\it SKA Memo 113}\footnote{Nijboer, Pandey-Pommier, \& de Bruyn; \url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

@@

% SKA
The low frequency component of the Square Kilometer Array (SKA), recently rebaselined to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending to 65~km, will operate from 50-350~MHz.\footnote{\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$), appropriate for lower mass ($\lesssim 10 M_J$) separately from the full band, which could be used to detect $\sim 30 M_J$ hot Jupiters.
We adopt the \cite{Rogers+Bowman2008} relation for $T_{\rm sys}$ as above, and assume $T_{\rm rx}\approx 60~\rm K$.
We caution that these assumptions are optimistic, and the time estimates could in reality be a factor of a few worse.

% VLA
For the low frequency receivers (28-80~MHz; 4-Band) in the \memoDS{Long Wavelength Array (LWA) and the?} Very Large Array (VLA), the receiver noise temperature $T_{\rm rx}\approx 260~\rm K$ \citep{Hicks2012} and is subdominant to $T_{\rm sky}$ by at least -6~dB.
Future upgrades to the VLA such as the LOw Band Observatory \citep[LOBO][]{Kassim2015IAU} are being considered, and would cover the full 30-470~MHz band.
We estimate the sensitivity of LOBO assuming the receiver noise extrapolates linearly from $T_{\rm rx}\approx 60~\rm K$ (internal VLA memo) in P-Band (230-470~MHz) to $T_{\rm rx}\approx 260~\rm K$ in 4-band, and consider just the 50-350~MHz portion of the full 30-470~MHz band.

>>>>

\newcommand{\Tcmb}{\mbox{$T_{\mbox{\tiny CMB}}$}}
\newcommand{\Tsky}{\mbox{$T_{\mbox{\tiny sky}}$}}
\newcommand{\Tsys}{\mbox{$T_{\mbox{\tiny sys}}$}}
\newcommand{\Trx}{\mbox{$T_{\mbox{\tiny rx}}$}}
\newcommand{\sigRMS}{\mbox{$\sigma_{\mbox{\tiny RMS}}$}}

@@

Another ingredient needed for modeling the radio emission from the interaction between the planetary magnetic field and the stellar wind is the density and velocity of the stellar wind. 
The number density of electrons in the stellar wind, $n$, can be expressed as
%%%%%%%%%% 
\begin{equation}
n = \frac{\dot M_\star}{4\pi a^2 m_p v}
\label{eq:n}
\end{equation}
%%%%%%%%%% 
where $\dot M_\star$ is the stellar mass loss rate, $a$ is the orbital distance from the star and $m_p$ is proton mass, and $v$ is the velocity of the stellar wind.

@@

We estimate the radio emission of exoplanets, $P_{\rm radio}$, simply by scaling the Jovian auroral radio emission, $P_{\rm radio,\,J}$, with the input energy from stellar wind, in the same manner as \citet{griesmeier2005,griesmeier2007a,griesmeier2007b}.
The scaling is based on the empirical/apparent good correlation between the radio emission intensity of Solar System planets and the input kinetic energy, $P_{\rm inp,\,k}$, or the magnetic energy of the solar wind, $P_{\rm inp,\,m}$, \citep[``radio Bode's law''; ][]{desch+kaiser1984}, i.e.,
%Namely, although the detail of the mechanism to generate planetary radio emission is not fully understood so far \memoYF{cite something?}, 
%observations indicate that the total radio emission, $P_{\rm radio}$ is proportional to the input kinetic energy, $P_{\rm k,\,inp}$, or the input magnetic energy, $P_{\rm m,\,inp}$ \citep[``radio Bode's law''; ][]{desch+kaiser1984}:
%%%
\begin{eqnarray}
P_{\rm radio} &\propto & P_{\rm inp} \\
P_{\rm inp,\,k} &=& n v^3 r_s ^2 \label{eq:Pinp_kin}, \\
P_{\rm inp,\,m} &=& v B_{\bot }^2 r_s ^2 \label{eq:Pinp_mag} \, ,
\end{eqnarray}
%%%
where $ B_{\bot }$ is the interstellar magnetic field perpendicular to the stellar wind flow, and $r_s$ is the radius of the magnetic stand-off point. 
%In the case of the Solar wind, the kinetic energy is dominant input energy over the magnetic one (the former is $\sim 400$ times larger than the latter).  
It is not clear from the observations of the Solar System planets which is the fundamental one, the correlation to the input kinetic energy, that to the magnetic energy, or both. 
In this paper, we assume that the radio emission is scaled with the input kinetic energy, i.e., $P_{\rm radio} \propto P_{\rm k,\,inp}$ and consider the possible effects of massive stellar wind.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{deluxetable*}{cccccc}
\tabletypesize{\footnotesize}
\tablecolumns{6}
\tablecaption{Detectability of Hot Jupiters with Current 
and Future Radio Telescopes.\label{tab:sens}}
\tablehead{
Instrument & Band  & \sigRMS\tablenotemark{a} & $\tau$\tablenotemark{b}  & Resolution & $\sigma_c$\tablenotemark{c}\\
           & (MHz) & ($\mu$Jy~hr$^{1/2}$)         &  (hr)                    & (\arcsec)  & ($\mu$Jy bm$^{-1}$)} 
\startdata
LOFAR-LBA  &  30-80  & 700   & 120000             & 8.3  & 42 \\ 
LOFAR-HBA  & 110-200 &  35   &    310             & 3.3  & 1.1 \\  
HERA       &  30-80  & 870   & 160000             & 1030 & 4.2$\times10^5$\\ 
GMRT       & 130-190 & 430   &  46000             & 16.5 & 56.9 \\ 
%LWA       &  30-80  & 58000 &  702127141         & 10.3 & 42.2 \\
%MWA       &  30-80  & 29000 &  176223211         & 1030 & 4.2$\times10^5$ \\
SKA-Low (part) & 50-200 & 7.8 &  12.6             & 7.6  & 9.7 \\
SKA-Low (full) & 50-350 & 2.2 &  1.0              & 4.8  & 1.5 \\
%VLA 4-Band& 30-80   & 3300  & 2312058            & 28.6 & 330 \\
VLA-LOBO   &  50-350 &  92   &  1800              & 8.6  & 10.4
\enddata
%\tablecomments{Time estimates to reach 2~$\mu$Jy RMS.}
\tablenotetext{a}{RMS noise in a 1-hour observation, computed for imaging of the full band reported 
in this table (Equation~\ref{eq:rms_sens}).}
\tablenotetext{b}{The integration time in hours required to reach an RMS of 2~$\mu$Jy (i.e.\ 5-$\sigma$ detection
of a typical source at 100~pc with 10~$\mu$Jy bm$^{-1}$ average flux).}
\tablenotetext{c}{Source confusion RMS contribution.}
\end{deluxetable*}
%
% HERA
% http://reionization.org/wp-content/uploads/2015/05/HERA0_imaging.pdf
%
% LOFAR
% https://www.astron.nl/~heald/test/sens.php; 
% 24 core stations, 16 remote, 8 international, 488 subbands
%
% GMRT
% http://gmrt.ncra.tifr.res.in/~astrosupp/obs_setup/sensitivity.html
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The primary limitation to detectability is a telescope's sensitivity, characterized by the root mean square (RMS) noise fluctuations in the sky and receiver.
The RMS noise \sigRMS\ in an interferometric observation with integration time $\tau$, bandwidth $\Delta \nu$, effective area $A_{\rm eff}$ ($\approx0.7 \pi (D/2)^2$ for interferometers comprising antennas with dish size $D$), and $N$ antennas is \citep[see e.g.][]{condon+ransom2016}
%%%
\begin{equation}
\sigRMS = \frac{2 k T_{\rm sys}}{A_{\rm eff} \, \sqrt{\Delta \nu \, \tau N (N-1)}} 
\left(\frac{\rm 10^{26} Jy}{\rm W m^{-2} Hz^{-1}}\right).
\label{eq:rms_sens}
\end{equation}
%%%
Here, \Tsys\ is the blackbody temperature equivalent of the total system noise, which is the sum of the instrument or receiver noise temperature \Trx\ and the noise contribution from the sky \Tsky.
We estimate \Tsky\ using the low frequency sky noise temperature fit from \cite{Rogers+Bowman2008}:
%%%
\begin{equation}
\Tsky = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + \Tcmb 
\end{equation}
%%%
where $T_{150} = 283.2~\rm K$ is the \Tsky\ at 150~MHz, $\beta=2.47$, and $\Tcmb = 2.73~\rm K$ is the contribution from the cosmic microwave background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors found it to be consistent with published measurements from 10-408~MHz. 
%The \Trx\ is the receiver noise temperature, which depends on the instruments and is described below.

@@

% LOFAR
The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their high and low bands, differing configurations, and antenna designs, which are called respectively the Low Band Antennas (LBA) and High Band Antennas (HBA).
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 100-240~MHz range \citep{vanHaarlem2013}.
We consider it as two separate instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and LOFAR-HBA's is at 150~MHz.
Given the strong frequency dependence of the effective aperture of a dipole antenna, and that the reciever noise temperature for LOFAR $\Trx \sim \Tsky$, we use the LOFAR Image Noise calculator\footnote{Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which is based on {\it SKA Memo 113}\footnote{Nijboer, Pandey-Pommier, \& de Bruyn; \url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

@@

% SKA
The low frequency component of the Square Kilometer Array (SKA), recently rebaselined to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending to 65~km, will operate from 50-350~MHz.\footnote{\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$), appropriate for lower mass ($\lesssim 10 M_J$) separately from the full band, which could be used to detect $\sim 30 M_J$ hot Jupiters.
We adopt the \cite{Rogers+Bowman2008} relation for \Tsys\ as above, and assume $\Trx \approx 60~\rm K$.
We caution that these assumptions are optimistic, and the time estimates could in reality be a factor of a few worse.

% VLA
For the low frequency receivers (28-80~MHz; 4-Band) in the \memoDS{Long Wavelength Array (LWA) and the?}\memoTM{No need to mention LWA here.  The receivers are the same in the VLA 4-band and the LWA, but the LWA is far too insensitive to consider here, as is VLA 4-band alone, so the reference just serves to place an upper limit on what \Trx\ will be for LOBO.} Very Large Array (VLA), the receiver noise temperature $\Trx \approx 260~\rm K$ \citep{Hicks2012} and is subdominant to \Tsky\ by at least -6~dB.
Future upgrades to the VLA such as the LOw Band Observatory \citep[LOBO][]{Kassim2015IAU} are being considered, and would cover the full 30-470~MHz band.
We estimate the sensitivity of LOBO assuming the receiver noise extrapolates linearly from $\Trx \approx 60~\rm K$ (internal VLA memo) in P-Band (230-470~MHz) to $\Trx \approx 260~\rm K$ in 4-band, and consider just the 50-350~MHz portion of the full 30-470~MHz band.
